Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-03-27 paper/plos.tex
address referees comments Tested everything on crystal.

* E-mail: Corresponding schuljef@physics.oregonstate.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust bipolar
  oscillations in a variety of cell shapes, recent experiments have
  shown that when the cells are mechanically deformed into wide,
  flattened out, irregular shapes, the spatial regularity of these
  oscillations breaks down. We employ widely used stochastic and
  deterministic models of the Min system to simulate cells with flattened
  shapes.  The deterministic model predicts strong bipolar
  oscillations, in contradiction with the experimentally observed
  behavior, while the stochastic model, which is based on the same
  reaction-diffusion equations, does predict spatially irregular
  oscillations as observed in experiment.  We further report
  simulations of symmetric but flattened cell shapes, and find that it
  is the flattening rather than the asymmetry of the cell shapes that
  causes the irregular oscillation behavior.

@@

\fixme{
  First referee:
  \begin{enumerate}
  \item \blue{Done} Add discussion of ``platform'' --- wrote our own code,
    publically available (cite github?).
  \item \blue{Done} Make 3D geometries available? Cite that the publically
    available code can generate them?
  \item Claims differential equations aren't needed.
  \item Claims wild-type movie-ish plot not needed.
  \item Points out problems with Huang 2003:
    \begin{enumerate}
    \item \blue{David.  I've added some sentences discussing the
      starry night in the context of the membrane diffusion.  I didn't
      say anything about a one half period allowing diffusion to smear
      an entire pole, because it seems to contradict the second part
      of this paragraph.  What do you think?} Lacks diffusion on the
      membrane. Supported by Meacci 2006, which experimentally tries
      to measure diffusion on membrane, has a tough time, but shows
      that definitely diffuses.
    \item Underestimates diffusion in cytoplasm by a factor of 5-6.5.
      Supported by Meacci 2006, who concludes a rate of 16 microns
      squared per sec for MinD and 10 microns squared per sec for
      MinE.
    \item Uses MinD-ADP -> MinD-ATP that is 3 times lower than
      experimental bound. Halatek uses ADP->ATP of 1/6 s-1. They cite
      Meacci 2006, who state that there is a lower bound on reference
      time of 3 seconds, so this constant has an upper bound of 1/3
      s-1.
    \item (Still debated) No evidence that DE recruits to membrane.
    \item Works in only a limited part of parameters phase space, and
      could not account for important things like
      temperature-independence.
    \end{enumerate}
  \item \blue{Done} Height should be bigger to compare with experiment.  0.4-0.6
    microns.  The 0.25 micron height of the cells used in this study
    does not represent the experimental settings well. The simulations
    are based on the flattened cell shapes reported from the
    experiments by Mannik et al 2009 and 2012. In these experimental
    studies, E. coli were shown to be able to grow in channels as low
    as 0.6 micron in hard silicon structures, and no more than 0.4
    micron (see Mannik 2009). Indeed, 0.25 micron channels were
    constructed with PDMS ceilings, which is a rather flexible polymer
    allowing bacteria to deform it to effectively higher
    channels. Thus the simulations would be more realistic if
    performed in the range of 0.4-0.6 micron thickness.

    \emph{If you can't make it grow in a .3 micron (rigid) silicon channel,
      then we and you observe it growing in a .25 micron flexible
      channel, it presumably must be able to grow because it deformed
      the channel to larger than .3 microns.}
  \item \blue{David} Starry night in the wild-type cells is a problem.
  \item Better explanation of correlation function (which is
    interesting!).
  \item Shape A is the interesting bit in our main figure.
  \item Would like more analysis of stochastic locations in different
    geometries.  Clusters?  Could add more geometries?
  \item Bold conclusions about flattening.  Talk about both width and
    thickness.  Compare Halatek 2012.  Wants systematically more
    geometries again.
  \item Is doubtful of our correlation plots relative to those of
    Fange and Elf 2006.
  \item Clarify (esp in conclusions) what we mean by flattening and
    irregular shape.  ``breaking the cylindrical symmetry and
    flattening the cells.'' versus ``breaking the mirror plane
    symmetry of the cross-section of a flattned cell.''
  \item[TODO suggestion] Switch to Halatek and Frey
  \item[TODO suggestion] Thoroughly explor shape parameter space
  \item[TODO suggestion] Check robustness of our conclusions with
    respect to changes in (reaction-diffusion?) parameters.
  \item Check order of figures and their references.
  \item Cite biological (non-computational) works e.g. Corbin 2008.
  \end{enumerate}
  Second referee:
  \begin{enumerate}
  \item Nitpicks about language of deterministic model versus
    stochastic model.  (Read carefully)
  \item \blue{David} Explain in greater detail the obviousness (maybe) that the
    deterministic approach doesn't have fluctuations.  RESPOND:
    chaotic systems do deterministically result in fluctuations.
    Explain this somewhere, and discuss triviality of the conclusion
    that deterministic is wrong because it does not exhibit these
    irregularities.
  \item Would like temporal averages over cell-cycle time scales.
  \item More discussion of how flattening leads to spatial
    irregularities.  Compare with large nearly spherical e. coli Shih
    2005.  See also Halatek and Frey 2012 for rotating oscillation
    axis due to destabilization of short modes.  How does this connect
    with our stability analysis?
  \item \blue{David} Should mention more recent revisions of the model.
  \item Fange and Elf do not use the same reaction-diffusion
    equations.
  \item Clarify statement about 1000 proteins and stochastic
    fluctuation.  Reword about a continuum description omitting
    information?
  \item \blue{Done} Do not call Huang parameters ``wild type values.''  They are
    fitted parameters.
  \item Add methods section for numerics (deterministic).
  \item \blue{Done} Add details of the linear stability analysis.
  \item \blue{David} Still need to fix periods on correlation plots.
    \blue{About this point - I updated the data and it seems the
      result is defferent from before - the stadium shapes seem to be
      temporally correlated for much longer compared with the shape A
      than before.}
  \end{enumerate}
}

@@

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are irregular, both temporally and spatially. Seen with
fluorescent microscopy, the MinD maximize in multiple locations within
the cell, in a seemingly random sequence. These experiments allow for
an opportunity to test MinD simulation models against more extreme
cases than have been seen thus far.  \fixme{\blue{(2)} An additional
  interesting question is whether the irregularity observed by Mannik
  \emph{et al.} indicates that the MinD system can become chaotic (as
  in e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this
irregularity is the result of a stochastic but nonchaotic process.}
\fixme{Jeff, please skim this chaos citation, which I cannot access
  from here.}

A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
approach with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%
\blue{(5)} Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately treat experimental findings
published after his 2003 paper
\cite{meacci2006mobility,loose2011min}. These models introduce slow
diffusion along the membrane (two to three orders of magnitude slower
than that within the cytoplasm), revise the process of MinD
recruitment to the membrane, and change some of the constants used in
the model.  In addition, Bonny \emph{et al.} assume an instantaneous
nucleotide exchange rate, and allows MinE to remain independently
attached to the membrane after dissasociating with
MinD~\cite{bonny2013membrane}.
%

%%  A number of studies have used an approach similar in
%% that they do not rely on the ability of MinD to move along the walls
%% and cluster~\cite{kruse2007experimentalist, meinhardt2001pattern,
%%   drew2005polymerization, fange2006noise, kerr2006division}, while
%% studies have been made as well of models which rely on MinD mobility
%% and attraction on the cell membrane~\cite{kruse2002dynamic,
%%   howard2005cellular}.

There are only on the order of 1000 MinD proteins in a given cell,
which means that stochastic fluctuation can be expected to play a
large role.  To treat this, there have been a number of variations of
the Huang 2003 model that stochastically simulate the same \fixme{similar}
reaction-diffusion equations~\cite{fange2006noise, kerr2006division}.
These studies largely confirm the results of Huang's deterministic
model when applied to the wild-type, pill-shaped phenotype.  However,
the stochastic models are slightly more successful in predicting
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's original 2003
model~\cite{huang2003dynamic}, which has been widely
used~\cite{hattne2005stochastic} \fixme{search for more} in both
deterministic and stochastic variants, to study flattened cells of
$0.4\micron$ thickness that are similar to those observed by Mannik
\emph{et al.}~\cite{mannik2009bacterial}.  \fixme{Some weak
  justification... it has been widely used.}

@@

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we follow Huang\cite{huang2003dynamic} and do not
deviate from the \blue{(8)} parameter values used in the cited work.

@@

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$,
respectively. \blue{(1,2)} We have chosen to write our own simulation
platform and membrane creation tools from scratch, instead of using
existing available software used in previous studies\fixme{cite
  previous studies} (we believe there is value in re-creating results
with alternative software).  Our simulations take place within a
three-dimensional Cartesian grid that has a grid spacing of
.05\micron, and our cell shapes are specified as the zero of an
analytic function which can generate the geometry presented in this
paper.  Our code is publically available on github.com~\fixme{cite
  github account in references? Yes!}

@@

\blue{(10)}  Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on a cylindrical model which shows an upper limit
on a steady state solution of a $2\micron$ half wavelength.  Cells
with dimensions longer than this spontaneously develop spatial
oscillations, while cells that are shorter relax into a homogeneous
steady state.  We have performed a similar linear stability analysis
on an infinite slab with a thickness equal to that of our flattened
cells ($0.40\micron$).  We first solve the model's five differential
equations for the steady state solution, under the constraint that the
total number MinD and MinE molecules matches our simulation.  We write
down as a matrix the linear response of the time derivatives of the
density to small, spatially harmonic density perturbations around the
steady-state density.  The system is stable at this wavelength
provided all eigenvalues of the matrix are negative, indicating that
all perturbations with this wavelength will decay.  We find the
stability limit by searching for the largest wavelength at which the
system is stable.  Through this method we arrive at an upper half
wavelength stability limit of $1.56\micron$ for our $0.40\micron$
flattened cells. As expected, when decreasing the lengths and widths
of our simulated flattened cells so that the longest distance across
the cell is less than this length, the cells stop exhibiting any
oscillatory behavior.  The deterministic model relaxes into a
motionless state and the stochastic model exhibits random fluctuations
without spatial oscillations.

@@

\blue{(5a),(7)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  This results
from the lack of inclusion of diffusion along the membrane in Huang's
model, which would spread out these recruitment clusters.  Experiments
have confirmed that diffusion along the membrane does occur, albeit
with a rate two orders of magnitude lower than that in the
cytoplasm~\cite{meacci2005min}.  \fixme{I would really like to see an
  estimate of the diffusivity width here.  A ``star'' from the starry
  night effect lasts about 10 seconds, so we could just figure out how
  far the protein would diffuse in 5 or 10 seconds.  If it comes out
  to close to 1 micron, we can then say that we do not expect this
  starry night effect to be observed in wild-type cells.}  \fixme{I'm
  not sure what the following sentence means.}  Even ignoring this
effect, however, it is clear from Fig.~\ref{image-p} that the
stochastic simulation results in a protein distribution that is less
spatially regular than the deterministic model predicts.  This
naturally leads to irregularity in the temporal periodicity of the
system, which we quantify by examining the temporal correlation
function of the total MinD found in two opposite polar regions.
%
This correlation function, which is displayed in Fig.~\ref{corr-pill},
is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

\begin{figure}
  \begin{center}
    \includegraphics[height=6.4cm]{../data/shape-randst/0_40-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}\\
    \includegraphics[height=6.4cm]{../data/shape-stad/0_40-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
  \end{center}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{A} (above) and the
    stadium \emph{A} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{A}, the
    stochastic model shows an oscillation period of 53.1 seconds and
    coherence time of 246 seconds, so that it takes roughly 4.6
    periods for the behavior to decohere.  For the stadium \emph{A},
    the same model shows an oscillation period of 51 seconds and
    coherence time of 323 seconds, so that it takes roughly 6.3
    periods for the behavior to decohere.}
  \label{fig:corr-pancake-A}
\end{figure}

\subsection*{Comparison of Experimental and Stadium Shapes}

As described in Section~\ref{sec:model-method-shapes}, we will focus
on simulation of four illustrative flattened cell shapes: two shapes
created to replicate those shapes observed experimentally by
Mannik \emph{et al.}~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two symmetrical stadium shapes
that have the identical aspect ratio, thickness and volume (stadium
\emph{A} and stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.
These stadium shapes enable us to distinguish between the effect of
flattening the cell and the cell shape's irregularity and asymmetry.
We note that shape \emph{B} is more irregular than shape \emph{A}, and
in particular features a prominent region of concavity on its
left-hand side.  For each of these shapes (in addition to the
wild-type pill shape discussed previously) we have simulated in excess
of 15000 seconds (over 4 hours) of evolution of the MinD system using
the deterministic version of Huang's model~\cite{huang2003dynamic},
and in excess of 100000 seconds (over 27 hours) using the stochastic
model.

>>>>

* E-mail: Corresponding schuljef@physics.oregonstate.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
  The dynamics of the Min-protein system help \emph{Escherichia coli}
  regulate the process of cell division by identifying the center of
  the cell.  While this system usually exhibits robust bipolar
  oscillations in a variety of cell shapes, recent experiments have
  shown that when the cells are mechanically deformed into wide,
  flattened out, irregular shapes, the spatial regularity of these
  oscillations breaks down. We employ widely used stochastic and
  deterministic models of the Min system to simulate cells with
  flattened shapes.  The deterministic model predicts strong bipolar
  oscillations, in contradiction with the experimentally observed
  behavior, while the stochastic model, which is based on the same
  reaction-diffusion equations, does predict spatially irregular
  oscillations as observed in experiment.  We further report
  simulations of symmetric but flattened cell shapes, and find that it
  is the flattening and lateral expansion rather than the asymmetry of
  the cell shapes that causes the irregular oscillation behavior.

@@

\fixme{
  First referee:
  \begin{enumerate}
  \item Done%% \blue{Done} Add discussion of ``platform'' --- wrote our own code,
    %% publically available (cite github?).
  \item Done
    %% \blue{Done} Make 3D geometries available? Cite that the publically
    %% available code can generate them?
  \item Done%%Claims differential equations aren't needed.
  \item Doen%%Claims wild-type movie-ish plot not needed.
  \item Points out problems with Huang 2003:
    \begin{enumerate}
    \item \blue{David.  I've added some sentences discussing the
      starry night effect and membrane diffusion blurring out the
      stars.  I have some confusion however about the diffusion
      lengths.}  Lacks diffusion on the membrane. Supported by Meacci
      2006, which experimentally tries to measure diffusion on
      membrane, has a tough time, but shows that definitely diffuses.
    \item \blue{David} Underestimates diffusion in cytoplasm by a factor of 5-6.5.
      Supported by Meacci 2006, who concludes a rate of 16 microns
      squared per sec for MinD and 10 microns squared per sec for
      MinE.
    \item \blue{David} Uses MinD-ADP -> MinD-ATP that is 3 times lower than
      experimental bound. Halatek uses ADP->ATP of 1/6 s-1. They cite
      Meacci 2006, who state that there is a lower bound on reference
      time of 3 seconds, so this constant has an upper bound of 1/3
      s-1.
    \item \blue{David} (Still debated) No evidence that DE recruits to membrane.
    \item Done%% Works in only a limited part of parameters phase space, and
      %% could not account for important things like
      %% temperature-independence.
    \end{enumerate}
  \item Done%% \blue{Done} Height should be bigger to compare with experiment.  0.4-0.6
    %% microns.  The 0.25 micron height of the cells used in this study
    %% does not represent the experimental settings well. The simulations
    %% are based on the flattened cell shapes reported from the
    %% experiments by Mannik et al 2009 and 2012. In these experimental
    %% studies, E. coli were shown to be able to grow in channels as low
    %% as 0.6 micron in hard silicon structures, and no more than 0.4
    %% micron (see Mannik 2009). Indeed, 0.25 micron channels were
    %% constructed with PDMS ceilings, which is a rather flexible polymer
    %% allowing bacteria to deform it to effectively higher
    %% channels. Thus the simulations would be more realistic if
    %% performed in the range of 0.4-0.6 micron thickness.

    %% \emph{If you can't make it grow in a .3 micron (rigid) silicon channel,
    %%   then we and you observe it growing in a .25 micron flexible
    %%   channel, it presumably must be able to grow because it deformed
    %%   the channel to larger than .3 microns.}
  \item Done%%\blue{David} Starry night in the wild-type cells is a problem.
  \item Done%%Better explanation of correlation function (which is
    %%interesting!).
  \item Done%%Shape A is the interesting bit in our main figure.
  \item Done%%Would like more analysis of stochastic locations in different
    %%geometries.  Clusters?  Could add more geometries?
  \item Done%%Bold conclusions about flattening.  Talk about both width and
    %% thickness.  Compare Halatek 2012.  Wants systematically more
    %% geometries again.
  \item Done%%Is doubtful of our correlation plots relative to those of
    %%Fange and Elf 2006.
  \item Done\blue{David. Read the refferee report with regard to this, and
    not sure what he wants...} Clarify (esp in conclusions) what we
    mean by flattening and irregular shape.  ``breaking the
    cylindrical symmetry and flattening the cells.'' versus ``breaking
    the mirror plane symmetry of the cross-section of a flattned
    cell.''
  \item Done%%[TODO suggestion] Switch to Halatek and Frey
  \item Done%%[TODO suggestion] Thoroughly explor shape parameter space
  \item Done%%[TODO suggestion] Check robustness of our conclusions with
    %%respect to changes in (reaction-diffusion?) parameters.
  \item Done%%Check order of figures and their references.
  \item Done%%%% \blue{David.  I cited the Corbin paper (although he's got a
    %% typo because Corbin 2008 is after the fange 2006 paper. I could
    %% cite other experimental papers, but the sentence in question
    %% specifically refers to changes made to the Huang model, and the
    %% papers cited are the main reasons that those changes have been
    %% made.)} Cite biological (non-computational) works, that the
    %% e.g. Corbin 2008.
  \end{enumerate}
  Second referee:
  \begin{enumerate}
  \item Done%%Nitpicks about language of deterministic model versus
    stochastic model.  (Read carefully)
  \item Done%%\blue{David} Explain in greater detail the obviousness (maybe) that the
    %% deterministic approach doesn't have fluctuations.  RESPOND:
    %% chaotic systems do deterministically result in fluctuations.
    %% Explain this somewhere, and discuss triviality of the conclusion
    %% that deterministic is wrong because it does not exhibit these
    %% irregularities.
  \item Would like temporal averages over cell-cycle time scales.
  \item More discussion of how flattening leads to spatial
    irregularities.  Compare with large nearly spherical e. coli Shih
    2005.  See also Halatek and Frey 2012 for rotating oscillation
    axis due to destabilization of short modes.  How does this connect
    with our stability analysis?
  \item Done%%\blue{David} Should mention more recent revisions of the model.
  \item Done%%\blue{David} Fange and Huang do not use the same reaction-diffusion
    %%equations.
  \item Done%%%% \blue{David.  I read the referee's note about this.  Not sure
    %% what he/she wants.}  Clarify statement about 1000 proteins and
    %% stochastic fluctuation.  Reword about a continuum description
    %% omitting information?
  \item Done%%%% \blue{Done} Do not call Huang parameters ``wild type values.''  They are
    %% fitted parameters.
  \item Done%%\blue{Done?  I think our additions to the Methods section
    %% addressed this?}  Add methods section for numerics
    %% (deterministic).
  \item Done%%\blue{Done} Add details of the linear stability analysis.
  \item Done%%\blue{David} Still need to fix periods on correlation plots.
    %% \blue{About this point - I updated the data and it seems the
    %%   result is defferent from before - the stadium shapes seem to be
    %%   temporally correlated for much longer compared with the shape A
    %%   than before.}
  \end{enumerate}
}

@@

However, Mannik \emph{et al.} have recently shown that there are
limitations to this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are irregular, both temporally and spatially. Seen with
fluorescent microscopy, the MinD maximize in multiple locations within
the cell, in a seemingly random sequence. These experiments allow for
an opportunity to test MinD simulation models against more extreme
cases than have been seen thus far.  An additional interesting
question is whether the irregularity observed by Mannik \emph{et al.}
indicates that the MinD system can become chaotic (as in
e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%
A number of models of the MinD protein system have been developed that
accurately describe its basic oscillatory nature.
%
Early models involved free proteins that affect each others' rates of
diffusion and membrane attachment, but do not combine into compound
states~\cite{meinhardt2001pattern}.  In 2003 Huang improved upon this
approach with a simple and very successful simulation model based on
MinD-MinE combination, ATPase hydrolysis, and MinD membrane attachment
that exhibits accurate MinD oscillations in cylindrical
cells~\cite{huang2003dynamic}. In this model cytoplasmic MinD is
recruited to the membrane by MinD that is already clustered there
(following observed non-linear attachment of MinD on the cell
membrane~\cite{hu2002dynamic,shih2002division}), and is stationary
once attached.
%

\blue{5a,b,c,d}Several models
\cite{fange2006noise,bonny2013membrane,halatek2012highly} modify
Huang's original model to more accurately reflect experimental
findings, which were published after his 2003
paper~\cite{meacci2006mobility,loose2011min}. These models introduce
diffusion along the membrane that is two to three orders of magnitude
slower than that within the cytoplasm, and increase the rate of
cytoplasmic diffusion by a factor of about five.  They also increase
or assume instantaneous~\cite{bonny2013membrane} the nucleotide
exchange rate.  In addition, they revise the process of MinD
recruitment to the membrane, prohibiting MinD from attaching to
MinD-MinE complexes there, behavior that is experimentally unresolved.
The model of Bonny \emph{et al.} allows MinE to remain independently
attached to the membrane after dissasociating with
MinD~\cite{bonny2013membrane}.
%

There are only on the order of 1000 MinD proteins in a given cell,
which is significantly less than the number of gridpoints used in our
simulations, suggesting that a continuum description will break down
and stochastic variation plays a significant role.  To treat this,
there have been a number of variations of the Huang 2003 model that
stochastically simulate the similar reaction-diffusion
equations~\cite{fange2006noise, kerr2006division}.  These studies
largely confirm the results of Huang's deterministic model when
applied to the wild-type, pill-shaped phenotype.  However, the
stochastic models are slightly more successful in predicting
experimentally observed oscillations in round cell
phenotypes~\cite{fange2006noise, huang2004min}, and they enable
prediction of fluctuations in the predicted
behavior~\cite{kruse2007experimentalist}.  Both deterministic and
stochastic models are widely used throughout systems
biology~\cite{lawson2013spatial, robb2014stochastic,
  oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
have unique advantages and limitations.  The deterministic approach
has an advantage in providing a simple prediction of average behavior,
while the stochastic approach enables prediction of fluctuations from
that mean, and reduces sensitivity to initial conditions.

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's original 2003
model~\cite{huang2003dynamic}, which has been widely
used~\cite{hattne2005stochastic,huang2004min,kerr2006division} in both
deterministic and stochastic variants, to study flattened cells of
$0.4\micron$ thickness that are similar to those observed by Mannik
\emph{et al.}~\cite{mannik2009bacterial}.

@@

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular size and shape on the protein
oscillations, so we use Huang's parameter
values~\cite{huang2003dynamic}.

@@

Huang's simulations use total MinD and MinE concentrations of
$1,000/\micron$ and $350/\micron$, respectively, in a cylindrical cell
of radius $0.5\micron$, and in our (non-cylindrical) cells we use the
same number of proteins per unit volume.  These concentration values
are $1273\micron^{-3}$ and $446\micron^{-3}$, respectively. We have
written our own simulation platform and membrane creation tools from
scratch, instead of using existing available software used in previous
studies~\cite{bonny2013membrane,fange2006noise,halatek2012highly,kerr2006division}
(we believe there is value in reproducing results with alternative
software).  Our simulations take place within a three-dimensional
Cartesian grid that has a grid spacing of .05\micron, and our cell
shapes are specified as the zero of an analytic function which can
generate the geometry presented in this paper.  Our code is publically
available on github~\cite{git}, including scripts that generate the
simulation data used in this paper.

@@

Huang \emph{et al.}~\cite{huang2003dynamic} has performed a linear
stability analysis on a cylindrical model which shows an upper limit
on a steady state solution of a $2\micron$ half wavelength.  Cells
with dimensions longer than this spontaneously develop spatial
oscillations, while cells that are shorter relax into a homogeneous
steady state.  We have performed a similar linear stability analysis
on an infinite slab with a thickness equal to that of our flattened
cells ($0.40\micron$).  We first solve the model's five differential
equations for the steady state solution, under the constraint that the
total number MinD and MinE molecules matches our simulation.  We write
down as a matrix the linear response of the time derivatives of the
density to small, spatially harmonic density perturbations around the
steady-state density.  The system is stable at this wavelength
provided all eigenvalues of the matrix are negative, indicating that
all perturbations with this wavelength will decay.  We find the
stability limit by searching for the largest wavelength at which the
system is stable.  Through this method we arrive at an upper half
wavelength stability limit of $1.56\micron$ for our $0.40\micron$
flattened cells. As expected, when decreasing the lengths and widths
of our simulated flattened cells so that the longest distance across
the cell is less than this length, the cells stop exhibiting any
oscillatory behavior.  The deterministic model relaxes into a
motionless state and the stochastic model exhibits random fluctuations
without spatial oscillations.

@@

\blue{(5a)} Both Fig.~\ref{image-p} and its associated movies (see
supplementary material), show that the stochastic model exhibits a
``starry night'' effect characterized by spatially fixed points of
protein density build up, as recruitment leads to clusters of MinD
forming on the wall and then subsequently dissipating.  This results
from the lack of inclusion of diffusion along the membrane in Huang's
model, which would allow these recruitment clusters to spread out.
Experiments have confirmed that diffusion along the membrane does
occur, albeit with a rate two orders of magnitude lower than that in
the cytoplasm~\cite{meacci2005min}.  The `stars' in the effect
typically last for around 10 seconds. The membrane-bound proteins will
travel a distance of around one micron during one of these periods of
time, enough to spread out across the polar section of the one micron
diameter wild-type pill cells.  As we see below, the flattened,
irregular cells are large enough so that the one micron blur would not
completely eliminate the `starry night' variation in maxima position
effect.

%% \fixme{calculating
  %% with $<x^2> = Dt$ and $D_{membrane} = .16\mu^2/s m$, which is just
  %% the experimental cytoplasmic value divided by 100, yields 1.2
  %% microns, which maybe is too high for what we're trying to say?
  %% Halatek uses $D_{membrane} = .013\mu^2/s m$, for which the same
  %% calculation would give .36 microns, which is better.  Halatek
  %% doesn't say why he uses a value that seems to be different than what
  %% Meacci claim to have found.  Halatek just says it's experimentally
  %% valid and cites their work.  What to do?}

We also analyze the temporal periodicity of the system, which we
quantify with temporal correlation function of the total MinD found in
two opposite polar regions.  This correlation function, which is
displayed in Fig.~\ref{corr-pill}, is given by
\begin{equation}
  C(\tau) \propto \int
  (N_{\textit{top}}(t) - \bar N_{\textit{top}})
  (N_{\textit{bottom}}(t+\tau) - \bar N_{\textit{bottom}})dt
\end{equation}
where $N_{\textit{top}}(t)$ and $N_{\textit{bottom}}(t)$ are the total
MinD proteins in the top and bottom thirds of a cell.  These plots
help us in studying the periods and the regularity and stability of
oscillations.  These curves are well fit with a simple decoherence
model, given by the equation
\begin{equation}
  C(\tau) = -\cos\left(\frac{2\pi\tau}{T}\right) e^{-\frac{\tau}{\tau_c}}
\end{equation}
where $T$ is the period and $\tau_c$ is the coherence time.  In the
case of the stochastic model the pill exhibits coherence times of
about 8 periods and in the case of the deterministic model oscillate
indefinitely with complete coherence, indicating that its behavior is
perfectly periodic.

@@

%% \begin{figure}
%%   \begin{center}
%%     \includegraphics[height=6.4cm]{../data/shape-randst/0_40-18_50-18_50-95_00-15_00-full_array/plots/correlation.pdf}\\
%%     \includegraphics[height=6.4cm]{../data/shape-stad/0_40-2_35-1_32-0_00-15_00-full_array/plots/correlation.pdf}
%%   \end{center}
%%   \caption{Temporal correlation function of the total MinD found in
%%     two opposite polar regions of the shape \emph{A} (above) and the
%%     stadium \emph{A} (below) cell shapes, shown against the
%%     correlation time.  Data for both the the determinisitic and
%%     stochastic models are shown, and the correlation functions are
%%     scaled to have the same initial value.  For shape \emph{A}, the
%%     stochastic model shows an oscillation period of 52 seconds and
%%     coherence time of 248 seconds, so that it takes roughly 4.8
%%     periods for the behavior to decohere.  For the stadium \emph{A},
%%     the same model shows an oscillation period of 48.4 seconds and
%%     coherence time of 634 seconds, so that it takes roughly 13
%%     periods for the behavior to decohere.}
%%   \label{fig:corr-pancake-A}
%% \end{figure}

\begin{figure}
  \begin{center}
    \includegraphics[width=8.7cm]{../data/shape-randst/0_40-18_60-28_60-94_00-15_00-full_array/plots/correlation.pdf}
    \includegraphics[width=8.7cm]{../data/shape-stad/0_40-2_92-1_18-0_00-15_00-full_array/plots/correlation.pdf}
  \end{center}
  \caption{Temporal correlation function of the total MinD found in
    two opposite polar regions of the shape \emph{B} (above) and the
    stadium \emph{B} (below) cell shapes, shown against the
    correlation time.  Data for both the the determinisitic and
    stochastic models are shown, and the correlation functions are
    scaled to have the same initial value.  For shape \emph{B}, the
    stochastic model shows an oscillation period of 58 seconds and
    coherence time of 289 seconds, so that it takes roughly 5 periods
    for the behavior to decohere.  For the stadium \emph{B}, the same
    model shows an oscillation period of 51 seconds and coherence time
    of 442 seconds, so that it takes roughly 8.6 periods for the
    behavior to decohere.}
  \label{fig:corr-B}
\end{figure}
